Generating Datasets about Building Designs \\Based on a Multilayer Building Ontology

### 2015-04-30 article-00-headers.tex
Updates from ShareLaTeX

\spnewtheorem{principle}{Principle}{\bfseries}{\rmfamily}

@@

@@

% ifcOWL layer names
\newcommand{\simple}{\textbf{Simple}}
\newcommand{\standard}{\textbf{Standard}}
\newcommand{\extended}{\textbf{Extended}}

\newcommand{\ifcowl}{\textbf{ifcOWL}}
\newcommand{\ifcsimple}{\ifcowl{}-\simple{}}
\newcommand{\ifcstandard}{\ifcowl{}-\standard{}}
\newcommand{\ifcextended}{\ifcowl{}-\extended{}}
\newcommand{\ifcrdf}{\textbf{ifcRDF}}

>>>>

%\spnewtheorem{principle}{Principle}{\bfseries}{\rmfamily}

\AtBeginDocument{
    \newtheoremstyle{principleStyle}
        {1.5\topsep} % Space above, default \topsep
        {0.5\topsep} % Space below, default \topsep
        {\itshape} % Body font
        {} % Indent amount, example: \parindent
        {\bfseries\itshape} % Theorem head font
        {.} % Punctuation after theorem head
        {.5em} % Space after theorem head
        {} % Theorem head spec (can be left empty, meaning `normal')

    \theoremstyle{principleStyle} \newtheorem{principle}{Principle}
}

@@

%
% THEOREMS
%
\usepackage{amsthm} % enables \newtheorem{}{}

@@

% ifcOWL layer names
\newcommand{\bname}[1]{\textnormal{\textbf{#1}}}
\newcommand{\simple}{\bname{Simple}}
\newcommand{\standard}{\bname{Standard}}
\newcommand{\extended}{\bname{Extended}}

\newcommand{\ifcowl}{\bname{ifcOWL}}
\newcommand{\ifcrdf}{\bname{ifcRDF}}
\newcommand{\ifcsimple}{\ifcowl{}-\simple{}}
\newcommand{\ifcstandard}{\ifcowl{}-\standard{}}
\newcommand{\ifcextended}{\ifcowl{}-\extended{}}

\newcommand{\ifcweb}{\bname{ifcWeb}}

### 2015-04-30 article-01-introduction.tex
Updates from ShareLaTeX

BIM models could provide a natural framework to organize and manage information about buildings. However, a big obstacle is the difficulty to access and utilize BIM models which have traditionally been shared shared through point-to-point file exchange. They cannot be used in an online, granular, and 
interlinked manner and provides a poor basis for open utilization of building information and the 
emergence of new, innovative applications.
To make BIM models accessible on the Web \cite{torma2014wobd}, we have developed the 
\name{WebIFC converter} to 
derive OWL ontologies from the IFC schema (\ifcowl{}), and RDF datasets from IFC files 
(\ifcrdf{}). The \ifcowl{} specifies concepts, relations, and constraints for \ifcrdf{} datasets that replicate the content of IFC files. 
All terms defined in \ifcowl{}  must thus be based on the corresponding constructs in 
the IFC schema.

@@

\vspace{2mm}
% \noindent
\begin{small}
\begin{tabular}{|l|l|l|}
 \hline
\textbf{Layer} & \textbf{Profiles} & \textbf{Included schema information} \\
\hline
 \hline
\name{ifcOWL-Simple} & OWL 2 EL & Schema meta data \\
& OWL 2 QL &  All type definitions and hierarchies \\
& OWL 2 RL & All entity type properties (names, range types) \\
\hline 
\name{ifcOWL-Standard} & OWL 2 RL & Schema meta data \\
& & All entity type keys \\
& & All entity type inverse properties \\
\hline
\name{ifcOWL-Extended} & OWL 2 RL & All cardinality constraints of properties \\
& & All cardinality constraints of list types \\
 \hline
\end{tabular}
\end{small}

@@

From the previous IFC-to-OWL/RDF conversions
\cite{beetz2005ontology,beetz2009ifcowl,pauwels2011interoperability} our approach differs since it
is based on OWL 2 profiles, produces a flexible multilayer ontology, and is specifically targeted to
real-life Linked Data applications. It uses stable URIs for entities, in contrast to the previous approaches
where the URIs were instable across models versions.

>>>>

BIM models could provide a natural framework to organize and manage information about buildings. However, a big obstacle is the difficulty to access and utilize BIM models which have traditionally been shared shared through point-to-point file exchange. They cannot be used in an online, granular, and 
interlinked manner and provides a poor basis for open utilization of building information and the 
emergence of new, innovative applications.
To make BIM models accessible on the Web \cite{torma2014wobd}, we have developed the 
\bname{WebIFC converter} to 
derive OWL ontologies from the IFC schema (\ifcowl{}), and RDF datasets from IFC files 
(\ifcrdf{}). The \ifcowl{} specifies concepts, relations, and constraints for \ifcrdf{} datasets that replicate the content of IFC files. 
All terms defined in \ifcowl{}  must thus be based on the corresponding constructs in 
the IFC schema.

@@

\vspace{0.5\topsep}
\noindent
\begin{scriptsize}
\begin{center}
    \def\arraystretch{1.2}
    \setlength{\tabcolsep}{6pt}
    \begin{tabularx}{0.95\textwidth}{|l|l|X|}
        \hline
            \textbf{Layer} & \textbf{Profile} & \textbf{Included schema information} \\
        \hline
            ifcOWL-Simple & OWL 2 EL or & Schema meta data plus: \\
            & OWL 2 QL or & \ \ - All type definitions and hierarchies \\
            & OWL 2 RL & \ \ - All entity type properties (names, range types) \\
        \hline 
            ifcOWL-Standard & OWL 2 RL & All above plus: \\
            & & \ \ - All entity type keys \\
            & & \ \ - All entity type inverse properties \\
        \hline
            ifcOWL-Extended & OWL 2 RL & All above plus: \\
            & & \ \ - All cardinality constraints \\
        \hline
    \end{tabularx}
\end{center}
\end{scriptsize}

@@

From the previous IFC-to-OWL/RDF conversions
\cite{beetz2005ontology,beetz2009ifcowl,pauwels2011interoperability} our approach differs since it
is based on OWL 2 profiles, produces a flexible multilayer ontology, applies \emph{very different methods of data\-type definitions} and is specifically targeted to
real-life Linked Data applications. It uses stable URIs for entities, in contrast to the previous approaches
where the URIs were instable across models versions.

### 2015-04-30 article-02-main.tex
Updates from ShareLaTeX

\section{Principles of Generating ifcOWL Ontologies And ifcRDF Datasets}
\label{sec:ifcOWL}
This section describes the derivation of the ifcOWL ontologies from the IFC specification. Since there are different options for the conversion, the choices are presented as conversion principles with justiciations.

\subsection{Common Conversion Principles}

\begin{principle}%[Simple OWL profiles]
\ifcsimple{} should be compatible with as simple OWL 2 profiles as possible (to support a maximum variety of reasoners and other tools). 
\end{principle}

The first layer \ifcsimple{} can be specified to be compatible with any OWL 2 DL profile: OWL 2 EL, OWL 2 QL or OWL 2 RL. The other layers \ifcstandard{} and \ifcextended{} are compatible with OWL 2 RL. OWL 2 RL needed because it -- unlike OWL 2 EL or QL -- supports such necessary things like enumerations of individuals, union classes, disjoint classes, and also anonymous individuals. \cite{w3c:owl2-profiles}

\begin{principle}%[No domain and range restrictions]
Property names should be kept simple and therefore domain and range constraints (specified by \name{rdfs:domain} and \name{rdfs:range}) should not be included in \ifcowl{} ontologies.
\end{principle}

In the IFC schema the property names are \emph{local} to entity classes while in RDF they are \emph{global}. 
In IFC the same property name can appear in multiple different classes. If the domain and range restrictions
were included in  \ifcowl{}, the property names would need to be qualified with entity class names. This 
would reduce the readability and usability of the \ifcowl{} significantly. In \ifcowl{} the domain and 
range restrictions would mainly be used to infer types, since there is no need to check type constraints 
like in programming languages \cite{w3c:owl-guide}. Moreover, the basic facilities given by them do not 
provide any direct way to indicate property restrictions that are local to a class \cite{w3c:rdf-schema}. 
Although it is possible to combine use \name{rdfs:domain} and \name{rdfs:range} with sub-property hierarchies, 
direct support for such declarations is provided by OWL constructs, e.g. \name{owl:\-onProperty}.

\begin{principle}%[Namespaces]
All constructs related to EXPRESS, STEP or IFC specifications must be belong to namespaces shown below with prefixes \name{expr:}, \name{step:} and \name{ifc:} accordingly (instead of \name{IFCXXX} there should be the name of IFC version, e.g. \name{IFC4\_ADD1}):
%\name{IFC2x3} or \name{IFC4}):
%At this moment the namespaces are defined as below (instead of \name{IFCXXX} there should be IFC versions, e.g. 
%\name{IFC2x3} or \name{IFC4}):

%\begin{lstlisting}[caption={Namespace definitions},label=lst:ifcOWL-namespaces]
\begin{lstlisting}
@prefix expr: <http://drumbeat.cs.hut.fi/owl/EXPRESS#>
@prefix step: <http://drumbeat.cs.hut.fi/owl/STEP#>
@prefix ifc: <http://drumbeat.cs.hut.fi/owl/IFCXXX#>
\end{lstlisting}
\end{principle}

\begin{principle}All definitions of functions, rules, type constraints and derived attributes specified in EXPRESS after keywords \name{FUNCTION}, \name{RULE}, \name{WHERE} and \name{RULE} accordingly should be ignored.
\end{principle}

\TODO{Explain why functions, rules, etc. should be ignored}

@@

\TODO{\textit{Write about EXPRESS in the Introduction section}:\\
The IFC specification is written as an EXPRESS schema. An EXPRESS schema defines series of data\-types, functions and rules by using EXPRESS data definition language specified in ISO10303-11 \cite{wiki:express,noauthor:ifc-guide}.}

@@

\subsubsection{Simple Datatypes} built in EXPRESS are: \name{STRING}, \name{BI\-NA\-RY}, \name{IN\-TE\-GER}, \name{REAL}, \name{NUM\-BER}, \name{BOO\-LEAN} and \name{LO\-GI\-CAL}. Types \name{STRING} and \name{BINARY} data\-types may be specified with a size parameter: \name{STRING (max\-Length\-In\-Bytes)} or \name{BINARY(size\-In\-Bits)}.

\begin{principle}%[Simple data\-types]
A simple data\-type (except \emph{BOOLEAN} and \emph{LOGICAL}) should be declared in \ifcowl{} as \name{owl:sameAs} of its most similar primitive XSD data\-type among all data\-types supported by any OWL 2 profile \cite{w3c:owl2-profiles} and preferred in Linked Data Platform \cite{w3c:ldp-best-practices}:

% \begin{lstlisting}[caption={Simple data types}, label=lst:ifcOWL-simple-types]
\begin{lstlisting}
expr:STRING, ifc:IfcGlobalUniqueId owl:sameAs xsd:string .
expr:BINARY owl:sameAs xsd:hexBinary .
expr:INTEGER owl:sameAs xsd:integer .
expr:REAL, expr:NUMBER owl:sameAs xsd:double .
\end{lstlisting}
\end{principle}

Type \name{LOGICAL} -- with its three possible values \name{TRUE}, \name{FALSE} and \name{UNKNOWN} -- is considered as an enumeration data\-type. The same to \name{BOOLEAN} because OWL 2 EL and QL do not support type \name{xsd:boolean} \cite{w3c:owl2-profiles}.

Statement `most similar XSD data\-type' means the type from XSD schema that conform to the value spaces of the EXPRESS simple data\-type better than any other type. For instance, type \name{xsd:double} is more similar to \name{REAL} than \name{xsd:decimal} as it is an IEEE 64-bit floating-point data\-type and supports scientific notation \cite{w3c:xsd,datapic:xsd}. When it is needed to refer to an EXPRRESS simple data\-type, we recommend to use its XSD equivalent type directly.

\begin{principle}%[Simple data\-types with sizes]
Size parameters of types \name{STRING} and \name{BINARY} (if any) should be ignored in layers \simple{} and \standard{} and should be declared obviously -- by means of OWL 2 data\-type facets -- in layer \extended{}, for instance:

% they are defined as equivalent to data\-types without size. But in layer \extended{}, the size parameter must be declared by using property \name{owl:onDatatype} and data\-type facets \name{xsd:maxLength} or \name{xsd:length} \cite{w3c:owl2-syntax,w3c:xmlschema11-2}, for instance:

% \begin{lstlisting}[caption={Size constraint of simple datatypes},label=lst:ifcOWL-simple-types-with-size]
\begin{lstlisting}
# only for Simple, Standard layers
  expr:STRING22 owl:sameAs xsd:string .
# only for Extended layer
  expr:STRING22 owl:equivalentClass [   a rdfs:Datatype ;
    owl:onDatatype  xsd:string; owl:withRestrictions ( [ xsd:length 22 ] ) ].
\end{lstlisting}
\end{principle}

@@

\subsubsection{Enumeration Datatypes} represent fixed sets of possible values.

\begin{principle}%[Enumeration data\-types]
Each enumeration data\-type should be defined in \ifcowl{} as an \name{owl:Class} which is a subclass of \name{expr:\-Enum\-Class}. Enumeration values must be defined as named individuals of the enumeration class but in different ways in different layers as shown below:

% \begin{lstlisting}[caption={Enumeration data\-types}, label=lst:ifcOWL-enum-types]
\begin{lstlisting}
# only for Simple layer
  ifc:IfcAssemblyPlaceEnum a owl:Class ; rdfs:subclassOf expr:EnumClass .
  ifc:SITE, ifc:FACTORY, ifc:NOTDEFINED rdf:type ifc:IfcAddressTypeEnum .
# only for Standard, Extended layers
  ifc:IfcAssemblyPlaceEnum a owl:Class ;  rdfs:subclassOf expr:EnumClass ;
    owl:oneOf ( ifc:SITE ifc:FACTORY ifc:NOTDEFINED ) .
\end{lstlisting}
\end{principle}

Two different approaches are used because enumerations with more than one individual or literal are not supported in OWL 2 EL or QL, like enumerations of literals in OWL 2 RL. The fact that some values -- for instance, \name{ifc:NOT\-DE\-FI\-NED} -- belong to different enumeration data\-types should not cause any problems since any enumeration value is always specified with an exact type.

As mentioned above, simple data\-types \name{BOOLEAN} and \name{LOGICAL} are considered as enumeration data\-types. But these types and their values must be defined in the EXPRESS-related namespace, e.g. \name{expr:LOGICAL}, \name{expr:TRUE}, \name{expr:UNKNOWN}.

@@

\subsubsection{Select Datatypes} represent unions of other data\-types. Particular members of a select data\-type can be entity, defined or select data\-types. For instance, type \name{Ifc\-Trim\-ming\-Se\-lect} is an union of entity data\-type \name{Ifc\-Car\-te\-sian\-Point} and defined data\-type \name{Ifc\-Pa\-ra\-me\-ter\-Va\-lue}.

\begin{principle}%[Select data\-types]
A select data\-type must be converted to an \name{owl:Class} which is a subclass of \name{expr:\-Select\-Class}. Its particular data\-types are defined as subclasses of the select class, but in different ways depended on the layer:
\end{principle}

@@

Two different approaches are used because property \name{owl:unionOf} is supported only in OWL 2 RL, but not in OWL 2 EL and QL \cite{w3c:owl2-profiles}.

@@

\begin{principle}%[Defined data\-types]
A defined data\-type based on a simple data\-type should be declared as an \name{owl:Class} which is a subclass of \name{expr:\-Defined\-Class} and has a single property \name{rdf:value} with that simple data\-type. A defined data\-type based on another defined data\-type should be declared as a subclass of the base defined data\-type, for instance:

%\begin{lstlisting}[caption={Defined data\-types}, label=lst:ifcOWL-defined-types]
\begin{lstlisting}
ifc:IfcLengthMeasure a owl:Class ; rdfs:subClassOf expr:DefinedClass ;
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty rdf:value;
                      owl:allValuesFrom xsd:double ] .
ifc:IfcPositiveLengthMeasure rdfs:subClassOf ifc:IfcLengthMeasure .
\end{lstlisting}
\end{principle}

Defined data\-types are defined as \name{owl:Class}es instead of \name{rdfs:Datatype}s because they can be merged with other \name{owl:Class}es for creating select data\-types. In addition, different defined data\-types wrapping the same \name{rdfs:Datatype} are not the same. For example, \name{IfcLengthMeasure} is different to \name{IfcMassMeasure} although they both are \name{REAL} and belong to select class \name{IfcMeasureValue}.

@@

\subsubsection{Entity Datatypes} are identical to classes in object-oriented programming languages. Each type may have one super\-type. An attribute is defined with a data\-type, cardinalities and can be marked as optional.

\begin{principle}%[Entity data\-types]
An entity data\-type should be converted to an \name{owl:\-Class} which is subclass of its superclass (if any) or \name{expr:\-Entity\-Class}. Entity attribute cardinalities can be only from 0 to 1, even if this attribute is a set of values. In this case the attribute type is an external aggregated data\-type, for example:

@@

# only for Standard & Extended layers
[] a owl:AllDisjointClasses ;
    owl:members ( ifc:IfcComplexProperty ifc:IfcSimpleProperty ) .
\end{lstlisting}
\end{principle}

@@

\begin{principle}Any aggregated type must be subclass of one of \name{expr:SetClass}, \name{expr:ListClass}, \name{expr:ArrayClass} and \name{expr:BagClass} which are subclasses of \name{expr:\-Collec\-tion\-Class}. The latter has one a nonfunctional property \name{expr:slot} with type \name{expr:Slot} and four functional properties: \name{expr:isOrdered}, \name{expr:item\-Type}, \name{expr:\-size}, \name{expr:\-start\-Index}, \name{expr:\-end\-Index}. Class \name{expr:Slot} has two functional properties \name{expr:item} (the actual collection item value), \name{expr:index} and also optional functional properties \name{expr:previous}, \name{expr:next}. Below is a fragment of definition of type \name{SET [2:?] OF IfcProfileDef}:
\end{principle}

@@

\subsubsection{Dataset Metadata}
%\label{subsec:ifcRDF-metadata}

@@

>>>>

\section{Principles of Generating ifcOWL Ontologies And ifcRDF Datasets}
\label{sec:ifcOWL}
This section describes the derivation of the ifcOWL ontologies from the IFC specification written as an EXPRESS schema. An EXPRESS schema defines series of data\-types, functions and rules by using EXPRESS data definition language specified in ISO10303-11 \cite{noauthor:ifc-guide}.

\subsection{Common Principles}

Layer \ifcsimple{} should be compatible with with any OWL 2 DL profile (EL, QL or RL) in order to support a maximum variety of reasoners and other tools. The other layers \standard{} and \extended{} must be compatible with OWL 2 RL which -- unlike OWL 2 EL or QL -- supports such useful things like enumerations of individuals, union classes and anonymous individuals \cite{motik2013owl}.

All \ifcowl{} layers must include all type definitions and hierarchies so that any \ifcrdf{} datasets converted from IFC data files could reuse that types without defining new ones.

Conversely, all function and rule definitions, type constraints and derived attributes specified in IFC schema after keywords \name{FUNCTION}, \name{RULE}, \name{WHERE} and \name{RULE} accordingly should not be translated into any \ifcowl{} ontology because that is either impossible, or can involve very complex constraints into \ifcowl{}.

Property names should be kept simple and therefore domain and range constraints (specified by \name{rdfs:domain} and \name{rdfs:range}) should not be included in \ifcowl{} ontologies. Firstly, there is no need to verify types or to infer types based on properties of \ifcowl{}. Secondly, in the IFC schema the property names are \emph{local} and the same property name can appear in multiple 
different classes. But in RDF property names are \emph{global} and the basic facilities given by domains and ranges do not provide any direct way to indicate property restrictions that are local to a class \cite{world2014rdf}. Finally, direct support for such declarations is provided by OWL constructs, e.g. \name{owl:\-onProperty}.

All constructs related to EXPRESS, STEP or IFC specifications must belong to separate namespaces with prefixes \name{expr:}, \name{step:} and \name{ifc:} accordingly (instead of \name{IFCXXX} there should be the name of IFC version, e.g. \name{IFC4\_ADD1}):

%\begin{lstlisting}[caption={Namespace definitions},label=lst:ifcOWL-namespaces]
\begin{lstlisting}
@prefix expr: <http://drumbeat.cs.hut.fi/owl/EXPRESS#>
@prefix step: <http://drumbeat.cs.hut.fi/owl/STEP#>
@prefix ifc: <http://drumbeat.cs.hut.fi/owl/IFCXXX#>
\end{lstlisting}

@@

@@

\subsubsection{Simple Datatypes} built in EXPRESS are: \name{STRING}, \name{BI\-NA\-RY}, \name{IN\-TE\-GER}, \name{REAL}, \name{NUM\-BER}, \name{BOO\-LEAN} and \name{LO\-GI\-CAL}. A simple data\-type (except \name{BOOLEAN} and \name{LOGICAL}) should be declared in \ifcowl{} as \name{owl:sameAs} of its most similar XSD data\-type among data\-types supported by all OWL 2 profiles \cite{motik2013owl} and preferred in Linked Data Platform \cite{ldp-best-practices}:

% \begin{lstlisting}[caption={Simple data types}, label=lst:ifcOWL-simple-types]
\begin{lstlisting}
expr:STRING, ifc:IfcGlobalUniqueId owl:sameAs xsd:string .
expr:BINARY owl:sameAs xsd:hexBinary .
expr:INTEGER owl:sameAs xsd:integer .
expr:REAL, expr:NUMBER owl:sameAs xsd:double .
\end{lstlisting}

Type \name{LOGICAL} -- with its three possible values \name{TRUE}, \name{FALSE} and \name{UNKNOWN} -- is considered as an enumeration data\-type. The same to \name{BOOLEAN} because OWL 2 EL and QL do not support type \name{xsd:boolean} \cite{motik2013owl}.

Type \name{xsd:double} is more similar to \name{REAL} than \name{xsd:decimal} as it is an IEEE 64-bit floating-point data\-type and supports scientific notation. When it is needed to refer to an EXPRRESS simple data\-type, we recommend to use its XSD equivalent type directly.

Types \name{STRING} and \name{BINARY} data\-types may be specified with a size parameter: \name{STRING(max\-Length\-In\-Bytes)} or \name{BINARY(size\-In\-Bits)}. In layers \simple{} and \standard{} the size should be ignored, while in layer \extended{} it should be declared:

% \begin{lstlisting}[caption={Size constraint of simple datatypes},label=lst:ifcOWL-simple-types-with-size]
\begin{lstlisting}
# only for Simple, Standard layers
  expr:STRING22 owl:sameAs xsd:string .
# only for Extended layer
  expr:STRING22 owl:equivalentClass [   a rdfs:Datatype ;
    owl:onDatatype  xsd:string; owl:withRestrictions ( [ xsd:length 22 ] ) ].
\end{lstlisting}

@@

\subsubsection{Enumeration Datatypes} represent fixed sets of possible string values. Each enumeration data\-type should be defined in \ifcowl{} as an \name{owl:Class} which is a subclass of \name{expr:\-Enum\-Class}\footnote{Hereinafter an object that has property \name{rdfs:subClassOf} is always an \name{owl:Class}}. Enumeration values must be defined as named individuals of the enumeration class but in different ways according to the layer:

% \begin{lstlisting}[caption={Enumeration data\-types}, label=lst:ifcOWL-enum-types]
\begin{lstlisting}
# only for Simple layer
  ifc:IfcAssemblyPlaceEnum a owl:Class ; rdfs:subclassOf expr:EnumClass .
  ifc:SITE, ifc:FACTORY, ifc:NOTDEFINED rdf:type ifc:IfcAddressTypeEnum .
# only for Standard, Extended layers
  ifc:IfcAssemblyPlaceEnum a owl:Class ;  rdfs:subclassOf expr:EnumClass ;
    owl:oneOf ( ifc:SITE ifc:FACTORY ifc:NOTDEFINED ) .
\end{lstlisting}

The enumeration values cannot be defined as literals because enumerations of literals is not supported by OWL 2 RL. Two different approaches are used because enumerations with more than one individual or literal are not supported in OWL 2 EL or QL. The fact that some values -- for instance, \name{ifc:NOT\-DE\-FI\-NED} -- belong to different enumeration data\-types should not cause any problems since any enumeration value is always specified with an exact type.

As mentioned, simple data\-types \name{BOOLEAN} and \name{LOGICAL} are considered as enumerations. But they and their values must be defined in the EXPRESS-related namespace, e.g. \name{expr:LOGICAL}, \name{expr:TRUE}, \name{expr:UNKNOWN}.

@@

\subsubsection{Select Datatypes} represent unions of other data\-types. Particular members of a select data\-type can be entity, defined or select data\-types. For instance, type \name{Ifc\-Trim\-ming\-Se\-lect} is an union of entity data\-type \name{Ifc\-Car\-te\-sian\-Point} and defined data\-type \name{Ifc\-Pa\-ra\-me\-ter\-Va\-lue}. A select data\-type must be converted to a subclass of \name{expr:\-Select\-Class}. Its particular data\-types are defined as subclasses of the select class, but in different ways according to the layer:

@@

Two different approaches are used because property \name{owl:unionOf} is not supported in OWL 2 EL and OWL 2 QL, but only in OWL 2 RL. \cite{motik2013owl}.

@@

A defined data\-type based on a simple data\-type should be declared as a subclass of \name{expr:\-Defined\-Class} and has a single property \name{rdf:value} with that simple data\-type. A defined data\-type based on another defined data\-type should be declared as a subclass of the base defined data\-type, for instance:

%\begin{lstlisting}[caption={Defined data\-types}, label=lst:ifcOWL-defined-types]
\begin{lstlisting}
ifc:IfcLengthMeasure a owl:Class ; rdfs:subClassOf expr:DefinedClass ;
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty rdf:value;
                      owl:allValuesFrom xsd:double ] .
ifc:IfcPositiveLengthMeasure rdfs:subClassOf ifc:IfcLengthMeasure .
\end{lstlisting}

Defined data\-types are declared as \name{owl:Class}es using boxing technique because they must be able to merge with other \name{owl:Class}es for creating select data\-types. In addition, values of defined data\-types must be qualified with type name. For example, value \name{"5.0"^^xsd:double} \TODO{}

for \name{IfcLengthMeasure} is different to \name{IfcMassMeasure} although they both are \name{REAL} and belong to select class \name{IfcMeasureValue}.

@@

\subsubsection{Entity Datatypes} are identical to classes in object-oriented programming languages. Each type may have one super\-type. An attribute is defined with a data\-type, cardinalities and can be marked as optional. An entity data\-type should be converted to an \name{owl:\-Class} which is subclass of its superclass (if any) or \name{expr:\-Entity\-Class}. Entity attribute cardinalities can be only from 0 to 1, even if this attribute is a set of values. In this case the attribute type is an external aggregated data\-type, for example:

@@

# only for Standard & Extended layers
[] a owl:AllDisjointClasses ;
    owl:members ( ifc:IfcComplexProperty ifc:IfcSimpleProperty ) .
\end{lstlisting}

@@

Any aggregated type in \ifcowl{} must be subclass of one of \name{expr:SetClass}, \name{expr:ListClass}, \name{expr:ArrayClass} and \name{expr:BagClass} which are subclasses of \name{expr:\-Collec\-tion\-Class}. The latter has one a nonfunctional property \name{expr:slot} with type \name{expr:Slot} and four functional properties: \name{expr:isOrdered}, \name{expr:item\-Type}, \name{expr:\-size}, \name{expr:\-start\-Index}, \name{expr:\-end\-Index}. Class \name{expr:Slot} has two functional properties \name{expr:item} (the actual collection item value), \name{expr:index} and also optional functional properties \name{expr:previous}, \name{expr:next}. Below is a fragment of definition of type \name{SET [2:?] OF IfcProfileDef}:

@@

\subsubsection{Header Section} ....
%\label{subsec:ifcRDF-metadata}

@@

\subsubsection{Data Section} ....
